---
title: "DEshort"
output: html_document
---

Differential expression analysis aims at distinguish the cluster of genes that are over- or under- represented in each sample. Comparing case and control samples aims at identifying which are the genes correlated to the disease.


### 0. importing data
Import the dataframe and filter unuseful information
```{r}
data <- read.table("R_steps/reads_gene_counts.txt", 
                   header = TRUE, 
                   sep = "\t")
colnames(data)
dim(data)
```

results in a table with 78334 obs for 21 variables (15 of which are our samples).
Then, filter for the columns corresponding to the samples and ignore the rest of the information. Specifically, remove the columns containing Chr, Start, End, Strand and Length.

```{r}
data <- data[,-2:-6]
colnames(data)
```

Create a matrix ready to feed into DESeq2. It has to be "numeric".
```{r}
rownames(data) <- data[,1]   # set first column as row names
data <- data[ , -1]          # remove the first column
```

## 5. Exploratory data analysis (DESeq2)
Install DESeq2
```
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("DESeq2")
```

Get the sample information into a table containing the sample name and to which group it belongs.
```{r}
sample_info <- read.delim(
  "R_steps/sample_info.txt",
  header = TRUE,
  sep = " ",
  stringsAsFactors = FALSE,
  check.names = FALSE
)
```

Now give the data to DEseq2 so that it can create a dds (DESeq Data Set that is created from the data matrix).

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
dds <- DESeq2::DESeqDataSetFromMatrix(countData = data,
                               colData = sample_info,
                               design= ~ Group)  # UNUSEFUL do not change df to matrix and then df again
```


Now remove the dependence of the variance on the mean by running `DESeq2::vst()` on dds. For visualization, we typically do not want to consider the experimental groups so set `blind = TRUE`.

```{r, echo=FALSE}
vsd <- DESeq2::vst(dds, blind = TRUE)
```

plot the PCA to check whether it is all right with the data and it is possible to distinguish in the 4 clusters where they belong to.

```{r}
DESeq2::plotPCA(vsd, intgroup = "Group")
```


## 6. Differential Expression analysis

```{r}
dss <- DESeq2::DESeq(dds)

dss

DESeq2::resultsNames(dss)
```

### check resultsNames and extract the comparisons of interest

To count how many are differentially expressed, check how many are not NA in the padj column and lower than the classic 0.05 threshold.

Check the number of differentially expressed genes in each comparison with *padj < 0.05*

- **WT case and DKO case**

```{r}
# res_WT_Case_KO_Case <- DESeq2::results(dss, contrast = list("Group_Lung_WT_Case_vs_Lung_DKO_Case"))   SAME AS:
res_WT_Case_KO_Case <- DESeq2::results(dss, contrast = c("Group", "Lung_WT_Case", "Lung_DKO_Case")) # extract comparison

WT_Case_KO_Case <- res_WT_Case_KO_Case[res_WT_Case_KO_Case$padj < 0.05 & ! is.na(res_WT_Case_KO_Case$padj), ] # filter for padj

sort_WT_KO_case <- WT_Case_KO_Case[order(WT_Case_KO_Case$padj), ] # order genes per lower padj

nrow(sort_WT_KO_case) # number of DE genes in this comparison
# head(sort_WT_KO_case) # fist 10 DE genes in this comparison

```

- **WT control vs WT case**
```{r}
res_WT_Control_WT_Case <- DESeq2::results(dss, contrast = c("Group", "Lung_WT_Case", "Lung_WT_Control")) # extract comparison

WT_Control_WT_Case <- res_WT_Control_WT_Case[res_WT_Control_WT_Case$padj < 0.05 & ! is.na(res_WT_Control_WT_Case$padj), ] # filter for padj

sort_WT <- WT_Control_WT_Case[order(WT_Control_WT_Case$padj), ] # order genes per lower padj

nrow(sort_WT) # number of DE genes in this comparison
# head(sort_WT) # fist 10 DE genes in this comparison
```

- **WT control vs DKO control**
```{r}
res_WT_Control_DKO_Control <- DESeq2::results(dss, contrast = c("Group", "Lung_WT_Control", "Lung_DKO_Control")) # extract comparison

WT_Control_DKO_Control <- res_WT_Control_DKO_Control[res_WT_Control_DKO_Control$padj < 0.05 & ! is.na(res_WT_Control_DKO_Control$padj), ] # filter for padj

sort_WT_DKO_d <- WT_Control_DKO_Control[order(WT_Control_DKO_Control$padj), ] # order genes per lower padj

nrow(sort_WT_DKO_d) # number of DE genes in this comparison
# head(sort_WT_DKO_d) # fist 10 DE genes in this comparison
```

- **DKO case vs DKO control**
```{r}
res_DKO_Control_DKO_Case <- DESeq2::results(dss, contrast = c("Group", "Lung_DKO_Control", "Lung_DKO_Case")) # extract comparison

DKO_Control_DKO_Case <- res_DKO_Control_DKO_Case[res_DKO_Control_DKO_Case$padj < 0.05 & ! is.na(res_DKO_Control_DKO_Case$padj), ] # filter for padj

sort_DKO <- DKO_Control_DKO_Case[order(DKO_Control_DKO_Case$padj), ] # order genes per lower padj

dim(sort_DKO) # number of DE genes in this comparison
# head(sort_DKO) # fist 10 DE genes in this comparison
```


### Heatmap
**Change the heatmap for each of the comparison you want to check**
Produce a heatmap to visualize the differences in the expression along all the samples.

```{r}
# 1. Extract VSD matrix
vsd_mat <- SummarizedExperiment::assay(vsd)

# 2. Select significant genes
res <- DESeq2::results(dss)
sig <- rownames(res)[which(res$padj < 0.05)] # sig padj < 0.05
sig_fc <- rownames(res)[which(res$padj < 0.05 & (res$log2FoldChange > 3 | res$log2FoldChange < -3))] # filter for FC
heat_mat <- vsd_mat[sig[1:100], ] # filtering for padj only
heat_mat_fc <- vsd_mat[sig_fc[1:100], ] # filtering for FC too

# 3. Build annotation
ann <- sample_info[, "Group", drop = FALSE]
rownames(ann) <- sample_info$Sample
ann <- ann[colnames(heat_mat), , drop = FALSE]


# 4. Plot heatmap
library(pheatmap)
  # padj filtering only
pheatmap(heat_mat,
         annotation_col = ann,
         show_rownames = FALSE,
         fontsize_col = 10,
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         clustering_method = "complete",
         scale = "row")
  # FC filtering
pheatmap(heat_mat_fc,
         annotation_col = ann,
         show_rownames = FALSE,
         fontsize_col = 10,
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         clustering_method = "complete",
         scale = "row")  # explain how scale = "row" or "column"
```



### Volcano plot


```{r}
# install.packages("devtools")
# devtools::install_github("kevinblighe/EnhancedVolcano")
library("EnhancedVolcano")
library("ggrepel")

EnhancedVolcano(res,
    lab = rownames(res),
    x = 'log2FoldChange',
    y = 'pvalue')
```


How many genes are differentially expressed (DE) in the pairwise comparison you selected (e.g. padj < 0.05)?
```{r}
paste0("There has been found ", length(sig), " DE genes in total.")

## WT Case against WT Control
res_WT_Control_WT_Case <- DESeq2::results(dss, contrast = list("Group_Lung_WT_Case_vs_Lung_DKO_Case",
                                                                 "Group_Lung_WT_Control_vs_Lung_DKO_Case"))

DE_WT_Control_vs_Case <- res_WT_Control_WT_Case[which(res_WT_Control_WT_Case$padj < 0.05 & !is.na(res_WT_Control_WT_Case$padj)), ]

paste0("The number of DE genes comparing WT control and cases is ", nrow(DE_WT_Control_vs_Case))
```
How many of the DE genes are up-regulated vs down-regulated?
```{r}
paste0("The number of DE genes upregulated for WT controls compared to cases are ", sum(DE_WT_Control_vs_Case$log2FoldChange > 0), " and ", length(DE_WT_Control_vs_Case$log2FoldChange) - sum(DE_WT_Control_vs_Case$log2FoldChange > 0), " downregulated.")

paste0("The number of DE genes having adjp < 0.05 and -1 < lFC < 1 is ", "FIND IT")
```

Based on the original publication, select 2-3 genes that are of particular interest and investigate their expression level. You could use, for example, the normalized counts (see DESeq2::counts) where the effect of between-sample differences in sequencing depth has been removed.



I can do this process for as many comparisons as I want. I need to read the paper to decide what to choose.

















--------------------------------------------------------------------------------


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 7. Overrepresentation analysis

Use the box plot to visualize fold change and significance (adj p-value) for the few selected genes for my report.


**Hypergeometric test** to check overrepresentation


The databases for overrepresentation test where you can get the pathway categorization are:
- GO
- KEGG
- Reactome
- Wikipaths

I would use GO and KEGG (more curated? check why don't go randomly)


Install clusterProfiler
```{r, message=FALSE, warning=FALSE}
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

if (!require("clusterProfiler", quietly = TRUE))
  BiocManager::install("clusterProfiler")
# the link to enrichplot is broken in bioconductor, install directly 
if (!require("enrichplot", quietly = TRUE))
  BiocManager::install("enrichplot", force = TRUE) 
# without updating the other packages (might create problems with dependencies)

library("clusterProfiler")
```

Download Bioconductor package with genome wide annotation for your species
```{r, message=FALSE, warning=FALSE}
if (!require("org.Mm.eg.db", quietly = TRUE))
  BiocManager::install("org.Mm.eg.db")
```


Call enrichGO
```{r}
# Select significant genes WT control and casees
sig <- rownames(res_WT_Control_WT_Case)[which(res_WT_Control_WT_Case$padj < 0.05)]

ego <- clusterProfiler::enrichGO(
  gene = sig, 
  universe = rownames(dss),
  OrgDb = org.Mm.eg.db,
  ont = "BP",
  keyType = "ENSEMBL",
  maxGSSize = 150,
  pvalueCutoff = 0.05
  )
```

- *gene = sig*
  
  sig should be a vector of ENSEMBL gene IDs that are significantly DE.
  
- *universe = rownames(mat)*

  All genes measured in the experiment. Since mat is your DESeq2 input matrix, its row names are all ENSEMBL IDs → perfect universe.
  
- *OrgDb = org.Mm.eg.db*

  Genome-wide mouse annotation package from Bioconductor.
  
- *ont = "BP"* or "MF", "CC", or "ALL"
  
  GO subontology:
  - BP: Biological Process
  - MF: Molecular Function
  - CC: Cellular Component

- *keyType = "ENSEMBL"*
  
  Because your gene IDs (row names) are ENSEMBL IDs like "ENSMUSG00000000001".

- *maxGSSize = 150*
  maximal size of genes annotated for testing


Work on the results of enrichGO
```{r}
dotplot(ego)
```

```{r}
dfego <- ego[grepl("interferon|ifn|ifit|oas", ego$Description, ignore.case=TRUE), ]

# find myeloid / granulocytes
mye_idx <- grepl("myeloid|granulocyte|neutrophil|il-1|interleukin-1|il-17|interleukin-17|immunoglobulin|antibody|cytotox|nk cell|natural killer", 
             ego$Description, ignore.case=TRUE)
```



```{r}
ego_df <- as.data.frame(ego)

keywords <- c(
  "granulocyte", "myeloid", "neutrophil",
  "immunoglobulin", "antibody", "cytotoxic", 
  "nk cell", "natural killer"
)

immune_ids <- ego_df[grepl(paste(keywords, collapse="|"), ego_df$Description, ignore.case = TRUE), ]

ego_immune <- ego
ego_immune@result <- ego@result[ego@result$ID %in% immune_ids$ID, ]

dotplot(ego_immune, showCategory = 20)
```







```
custom_sets <- list(
  Granulocyte = c("Ly6g","S100a8","S100a9","Mpo"),
  NK_cells    = c("Nkg7","Gzmb","Prf1"),
  Immunoglobulin = c("Ighm","Ighg1","Igkc","Iglc1")
)
```


