---
title: "DE analysis"
output: html_document
---

Differential expression analysis aims at distinguish the cluster of genes that are over- or under- represented in each sample. Comparing case and control samples aims at identifying which are the genes correlated to the disease.


### 0. importing data
Import the dataframe and filter unuseful information
```{r}
data <- read.table("reads_gene_counts.txt", 
                   header = TRUE, 
                   sep = "\t")
colnames(data)
dim(data)
```

results in a table with 78334 obs for 21 variables (15 of which are our samples).
Then, filter for the columns corresponding to the samples and ignore the rest of the information. Specifically, remove the columns containing Chr, Start, End, Strand and Length.

```{r}
data <- data[,-2:-6]
colnames(data)
```

Create a matrix ready to feed into DESeq2. It has to be "numeric".
```{r}
rownames(data) <- data[,1]   # set first column as row names
data <- data[ , -1]          # remove the first column
```

## 5. Exploratory data analysis (DESeq2)
Install DESeq2
```
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("DESeq2")
```

Get the sample information into a table containing the sample name and to which group it belongs.
```{r}
sample_info <- read.delim(
  "sample_info.txt",
  header = TRUE,
  sep = " ",
  stringsAsFactors = FALSE,
  check.names = FALSE
)
```

Now give the data to DEseq2 so that it can create a dds (DESeq Data Set that is created from the data matrix).

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
dds <- DESeq2::DESeqDataSetFromMatrix(countData = data,
                               colData = sample_info,
                               design= ~ Group)  # UNUSEFUL do not change df to matrix and then df again
```


Now remove the dependence of the variance on the mean by running `DESeq2::vst()` on dds. For visualization, we typically do not want to consider the experimental groups so set `blind = TRUE`.

```{r, echo=FALSE}
vsd <- DESeq2::vst(dds, blind = TRUE)
```

plot the PCA to check whether it is all right with the data and it is possible to distinguish in the 4 clusters where they belong to.

```{r}
DESeq2::plotPCA(vsd, intgroup = "Group")
```


DESeq2::plotPCA() does NOT return the PCA object, it only plots it.
But the PCA is computed on the top variable genes of the variance-stabilized matrix (vsd), so from there we can extract the information on the top variable genes that explain most of the variation for this database (loadings).
```{r}
# Extract the assay matrix from VSD
mat_vsd <- SummarizedExperiment::assay(vsd)

# Select the top variable genes
library("genefilter")
rv <- rowVars(mat_vsd)    # BiocManager::install("genefilter")
select <- order(rv, decreasing = TRUE)[1:500]

# Transpose and run PCA
pca <- prcomp(t(mat_vsd[select, ]))

# Inspect the PCA results
## Variance explained by PCs
summary(pca)

# Visualize the variance explained (useful to see how many PCs matter)
barplot( (pca$sdev^2 / sum(pca$sdev^2))[1:7],
         main = "Variance explained by first 7 PCs",
         xlab = "Principal component", ylab = "Proportion of variance" )

## Loadings (genes contributing to PCs)
loadings <- pca$rotation   # genes x PCs
head(loadings[, 1:5])      # first 5 PCs
```

## 6. Differential Expression analysis

```{r}
dss <- DESeq2::DESeq(dds)

dss

DESeq2::resultsNames(dss)
```

### check resultsNames and extract the comparisons of interest

To count how many are differentially expressed, check how many are not NA in the padj column and lower than the classic 0.05 threshold.

Check the number of differentially expressed genes in each comparison with *padj < 0.05*

- **WT case and DKO case**

```{r}
# res_WT_Case_DKO_Case <- DESeq2::results(dss, contrast = list("Group_Lung_WT_Case_vs_Lung_DKO_Case"))   SAME AS:
res_WT_Case_DKO_Case <- DESeq2::results(dss, contrast = c("Group", "Lung_WT_Case", "Lung_DKO_Case"))

WT_Case_DKO_Case <- res_WT_Case_DKO_Case[res_WT_Case_DKO_Case$padj < 0.05 & ! is.na(res_WT_Case_DKO_Case$padj), ]

sort_WT_DKO_case <- WT_Case_DKO_Case[order(WT_Case_DKO_Case$padj), ]

nrow(sort_WT_DKO_case)
head(sort_WT_DKO_case)

```

- **WT control vs WT case**
```{r}
res_WT_Control_WT_Case <- DESeq2::results(dss, contrast = c("Group", "Lung_WT_Case", "Lung_WT_Control"))

WT_Control_WT_Case <- res_WT_Control_WT_Case[res_WT_Control_WT_Case$padj < 0.05 & ! is.na(res_WT_Control_WT_Case$padj), ]

sort_WT <- WT_Control_WT_Case[order(WT_Control_WT_Case$padj), ]

nrow(sort_WT)
head(sort_WT)
```

- **WT control vs DKO control**
```{r}
res_WT_Control_DKO_Control <- DESeq2::results(dss, contrast = c("Group", "Lung_WT_Control", "Lung_DKO_Control"))

WT_Control_DKO_Control <- res_WT_Control_DKO_Control[res_WT_Control_DKO_Control$padj < 0.05 & ! is.na(res_WT_Control_DKO_Control$padj), ]

sort_WT_DKO_contr <- WT_Control_DKO_Control[order(WT_Control_DKO_Control$padj), ]

nrow(sort_WT_DKO_contr)
head(sort_WT_DKO_contr)
```

- **DKO case vs DKO control**
```{r}
res_DKO_Control_DKO_Case <- DESeq2::results(dss, contrast = c("Group", "Lung_DKO_Control", "Lung_DKO_Case"))

DKO_Control_DKO_Case <- res_DKO_Control_DKO_Case[res_DKO_Control_DKO_Case$padj < 0.05 & ! is.na(res_DKO_Control_DKO_Case$padj), ]

sort_DKO <- DKO_Control_DKO_Case[order(DKO_Control_DKO_Case$padj), ]

dim(sort_DKO)
head(sort_DKO)
```


### Heatmap

Produce a heatmap to visualize the differences in the expression along all the samples.

**DKO control vs cases**
```{r}
# 1. Extract VSD matrix
vsd_mat <- SummarizedExperiment::assay(vsd)

# 2. Select significant genes for DKO comparison
filt_DKO <- DKO_Control_DKO_Case[(DKO_Control_DKO_Case$log2FoldChange > 3 | DKO_Control_DKO_Case$log2FoldChange < -3), ]
filt_DKO <- filt_DKO[order(filt_DKO$padj),]
filt_DKO <- filt_DKO[1:50,]
heat_DKO <- vsd_mat[rownames(filt_DKO),]


# 3. Build annotation for DKO comparison
ann <- sample_info[, "Group", drop = FALSE]
rownames(ann) <- sample_info$Sample
ann <- ann[colnames(heat_DKO), , drop = FALSE]


# 4. Plot heatmap
library(pheatmap)
  # padj filtering only
pheatmap(heat_DKO,
         annotation_col = ann,
         show_rownames = TRUE,
         fontsize_col = 10,
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         clustering_method = "complete",
         scale = "row")
```
Repeat steps from 2 to 4 for all the other comparisons

**WT vs DKO control**
```{r}
# 2. Select significant genes for DKO comparison
filt_Control <- WT_Control_DKO_Control[(WT_Control_DKO_Control$log2FoldChange > 3 | WT_Control_DKO_Control$log2FoldChange < -3), ]
filt_Control <- filt_Control[order(filt_Control$padj),]
filt_Control <- filt_Control[1:50,]
heat <- vsd_mat[rownames(filt_Control),]


# 3. Build annotation for DKO comparison
ann <- ann[colnames(heat_DKO), , drop = FALSE]


# 4. Plot heatmap
library(pheatmap)
  # padj filtering only
pheatmap(heat,
         annotation_col = ann,
         show_rownames = TRUE,
         fontsize_col = 10,
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         clustering_method = "complete",
         scale = "row")
```

**WT control vs case **
```{r}
# 2. Select significant genes for DKO comparison
filt_WT <- WT_Control_WT_Case[(WT_Control_WT_Case$log2FoldChange > 3 | WT_Control_WT_Case$log2FoldChange < -3), ]
filt_WT <- filt_WT[order(filt_WT$padj),]
filt_WT <- filt_WT[1:50,]
heat <- vsd_mat[rownames(filt_WT),]


# 3. Build annotation for DKO comparison
ann <- ann[colnames(heat_DKO), , drop = FALSE]


# 4. Plot heatmap
library(pheatmap)
  # padj filtering only
pheatmap(heat,
         annotation_col = ann,
         show_rownames = TRUE,
         fontsize_col = 10,
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         clustering_method = "complete",
         scale = "row")
```


**WT vs DKO case**
```{r}
# 2. Select significant genes for DKO comparison
filt_Case <- WT_Case_DKO_Case[(WT_Case_DKO_Case$log2FoldChange > 3 | WT_Case_DKO_Case$log2FoldChange < -3), ]
filt_Case <- filt_Case[order(filt_Case$padj),]
filt_Case <- filt_Case[1:50,]
heat <- vsd_mat[rownames(filt_Case),]


# 3. Build annotation for DKO comparison
ann <- ann[colnames(heat_DKO), , drop = FALSE]


# 4. Plot heatmap
library(pheatmap)
  # padj filtering only
pheatmap(heat,
         annotation_col = ann,
         show_rownames = TRUE,
         fontsize_col = 10,
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         clustering_method = "complete",
         scale = "row")
```

### Volcano plot


```{r}
# install.packages("devtools")
# devtools::install_github("kevinblighe/EnhancedVolcano")
library("EnhancedVolcano")
library("ggrepel")

EnhancedVolcano(filt_Case,
    lab = rownames(filt_Case),
    x = 'log2FoldChange',
    y = 'pvalue')
```


**MAKE THE FOLLOWING A TABLE IN THE REPORT**

How many genes are differentially expressed (DE) in the pairwise comparison you selected (e.g. padj < 0.05)? 
```{r}
paste0("There has been found ", nrow(WT_Control_WT_Case), " DE genes in WT control vs case comparison having padj < 0.05.")

paste0("There has been found ", nrow(WT_Case_DKO_Case), " DE genes in WT vs DKO case comparison having padj < 0.05.")

paste0("There has been found ", nrow(WT_Control_DKO_Control), " DE genes in WT vs DKO control comparison having padj < 0.05.")

paste0("There has been found ", nrow(DKO_Control_DKO_Case), " DE genes in DKO control vs case comparison having padj < 0.05.")
```
How many of the DE genes are up-regulated vs down-regulated meaningfully (padj < 0.05)?
```{r}
paste0("For WT controls vs cases are ", sum(WT_Control_WT_Case$log2FoldChange > 0), " upregulated and ", length(WT_Control_WT_Case$log2FoldChange) - sum(WT_Control_WT_Case$log2FoldChange > 0), " downregulated.")

paste0("For WT vs DKO case are ", sum(WT_Case_DKO_Case$log2FoldChange > 0), " upregulated and ", length(WT_Case_DKO_Case$log2FoldChange) - sum(WT_Case_DKO_Case$log2FoldChange > 0), " downregulated.")

paste0("For WT vs DKO control are ", sum(WT_Control_DKO_Control$log2FoldChange > 0), " upregulated and ", length(WT_Control_DKO_Control$log2FoldChange) - sum(WT_Control_DKO_Control$log2FoldChange > 0), " downregulated.")

paste0("For DKO control vs case are ", sum(DKO_Control_DKO_Case$log2FoldChange > 0), " upregulated and ", length(DKO_Control_DKO_Case$log2FoldChange) - sum(DKO_Control_DKO_Case$log2FoldChange > 0), " downregulated.")
```

Based on the original publication, select 2-3 genes that are of particular interest and investigate their expression level. You could use, for example, the normalized counts (see DESeq2::counts) where the effect of between-sample differences in sequencing depth has been removed.



















--------------------------------------------------------------------------------


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 7. Overrepresentation analysis

Use the box plot to visualize fold change and significance (adj p-value) for the few selected genes for my report.


**Hypergeometric test** to check overrepresentation


The databases for overrepresentation test where you can get the pathway categorization are:
- GO
- KEGG
- Reactome
- Wikipaths

I would use GO and KEGG (more curated? check why don't go randomly)


Install clusterProfiler
```{r, message=FALSE, warning=FALSE}
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

if (!require("clusterProfiler", quietly = TRUE))
  BiocManager::install("clusterProfiler")
# the link to enrichplot is broken in bioconductor, install directly 
if (!require("enrichplot", quietly = TRUE))
  BiocManager::install("enrichplot", force = TRUE) 
# without updating the other packages (might create problems with dependencies)

library("clusterProfiler")
```

Download Bioconductor package with genome wide annotation for your species
```{r, message=FALSE, warning=FALSE}
if (!require("org.Mm.eg.db", quietly = TRUE))
  BiocManager::install("org.Mm.eg.db")
```


Call enrichGO
```{r}
# Select significant genes WT control and casees
sig <- rownames(res_WT_Control_WT_Case)[which(res_WT_Control_WT_Case$padj < 0.05)]

ego <- clusterProfiler::enrichGO(
  gene = sig, 
  universe = rownames(mat),
  OrgDb = org.Mm.eg.db,
  ont = "BP",
  keyType = "ENSEMBL",
  maxGSSize = 150,
  pvalueCutoff = 0.05
  )
```

- *gene = sig*
  
  sig should be a vector of ENSEMBL gene IDs that are significantly DE.
  
- *universe = rownames(mat)*

  All genes measured in the experiment. Since mat is your DESeq2 input matrix, its row names are all ENSEMBL IDs → perfect universe.
  
- *OrgDb = org.Mm.eg.db*

  Genome-wide mouse annotation package from Bioconductor.
  
- *ont = "BP"* or "MF", "CC", or "ALL"
  
  GO subontology:
  - BP: Biological Process
  - MF: Molecular Function
  - CC: Cellular Component

- *keyType = "ENSEMBL"*
  
  Because your gene IDs (row names) are ENSEMBL IDs like "ENSMUSG00000000001".

- *maxGSSize = 150*
  maximal size of genes annotated for testing


Work on the results of enrichGO
```{r}
dotplot(ego)
```

```{r}
dfego <- ego[grepl("interferon|ifn|ifit|oas", ego$Description, ignore.case=TRUE), ]

# find myeloid / granulocytes
mye_idx <- grepl("myeloid|granulocyte|neutrophil|il-1|interleukin-1|il-17|interleukin-17|immunoglobulin|antibody|cytotox|nk cell|natural killer", 
             ego$Description, ignore.case=TRUE)
```



```{r}
ego_df <- as.data.frame(ego)

keywords <- c(
  "granulocyte", "myeloid", "neutrophil",
  "immunoglobulin", "antibody", "cytotoxic", 
  "nk cell", "natural killer"
)

immune_ids <- ego_df[grepl(paste(keywords, collapse="|"), ego_df$Description, ignore.case = TRUE), ]

ego_immune <- ego
ego_immune@result <- ego@result[ego@result$ID %in% immune_ids$ID, ]

dotplot(ego_immune, showCategory = 20)
```







```
custom_sets <- list(
  Granulocyte = c("Ly6g","S100a8","S100a9","Mpo"),
  NK_cells    = c("Nkg7","Gzmb","Prf1"),
  Immunoglobulin = c("Ighm","Ighg1","Igkc","Iglc1")
)
```


