---
title: "Overrepresentation analysis"
author: "Isaac Ambrogetti"
date: "2025-12-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 7. Overrepresentation analysis

Use the box plot to visualize fold change and significance (adj p-value) for the few selected genes for my report.


**Hypergeometric test** to check overrepresentation


The databases for overrepresentation test where you can get the pathway categorization are:
- GO
- KEGG
- Reactome
- Wikipaths

I would use GO and KEGG (more curated? check why don't go randomly)


Install clusterProfiler
```{r, message=FALSE, warning=FALSE}
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

if (!require("clusterProfiler", quietly = TRUE))
  BiocManager::install("clusterProfiler")
# the link to enrichplot is broken in bioconductor, install directly 
if (!require("enrichplot", quietly = TRUE))
  BiocManager::install("enrichplot", force = TRUE) 
# without updating the other packages (might create problems with dependencies)

library("clusterProfiler")
```

Download Bioconductor package with genome wide annotation for your species
```{r, message=FALSE, warning=FALSE}
if (!require("org.Mm.eg.db", quietly = TRUE))
  BiocManager::install("org.Mm.eg.db")
```


Call enrichGO
```{r}
# Select significant genes WT control and casees
sig <- rownames(res_WT_Control_WT_Case)[which(res_WT_Control_WT_Case$padj < 0.05)]

ego <- clusterProfiler::enrichGO(
  gene = sig, 
  universe = rownames(mat),
  OrgDb = org.Mm.eg.db,
  ont = "BP",
  keyType = "ENSEMBL",
  maxGSSize = 150,
  pvalueCutoff = 0.05
  )
```

- *gene = sig*
  
  sig should be a vector of ENSEMBL gene IDs that are significantly DE.
  
- *universe = rownames(mat)*

  All genes measured in the experiment. Since mat is your DESeq2 input matrix, its row names are all ENSEMBL IDs → perfect universe.
  
- *OrgDb = org.Mm.eg.db*

  Genome-wide mouse annotation package from Bioconductor.
  
- *ont = "BP"* or "MF", "CC", or "ALL"
  
  GO subontology:
  - BP: Biological Process
  - MF: Molecular Function
  - CC: Cellular Component

- *keyType = "ENSEMBL"*
  
  Because your gene IDs (row names) are ENSEMBL IDs like "ENSMUSG00000000001".

- *maxGSSize = 150*
  maximal size of genes annotated for testing


Work on the results of enrichGO
```{r}
dotplot(ego)
```

```{r}
dfego <- ego[grepl("interferon|ifn|ifit|oas", ego$Description, ignore.case=TRUE), ]

# find myeloid / granulocytes
mye_idx <- grepl("myeloid|granulocyte|neutrophil|il-1|interleukin-1|il-17|interleukin-17|immunoglobulin|antibody|cytotox|nk cell|natural killer", 
             ego$Description, ignore.case=TRUE)
```



```{r}
ego_df <- as.data.frame(ego)

keywords <- c(
  "granulocyte", "myeloid", "neutrophil",
  "immunoglobulin", "antibody", "cytotoxic", 
  "nk cell", "natural killer"
)

immune_ids <- ego_df[grepl(paste(keywords, collapse="|"), ego_df$Description, ignore.case = TRUE), ]

ego_immune <- ego
ego_immune@result <- ego@result[ego@result$ID %in% immune_ids$ID, ]

dotplot(ego_immune, showCategory = 20)
```












```{r}
keywords <- append(BP_one, BP_two)

ego_plot <- function(ego, keywords, name){
  ego_fil <- ego
  
  # filter for keywords of type I and type II TFN present in the enrichment of the comparison chosen
  ego_fil@result <- ego@result[ego@result$Description %in% keywords, ]
  
  dotplot(ego_fil, showCategory = 10, title = name) + theme(axis.text.y = element_text(size = 9)) + coord_cartesian(xlim = c(0.005, 0.013))
}

ego_plot(ego_WT, keywords, "WT controls vs cases")
ego_plot(ego_DKO, keywords, "DKO controls vs cases")
ego_plot(ego_Case, keywords, "WT vs DKO cases")
```